#440 Reproducibility and Statistics Assignment Homework

<!-- This is an Rmd comment. -->
<!--
You can make comments several lines long by inserting
the text in between the the start and the end of the 
comment.

Comments are where I'll tell you what to do. You should
use the space directly below the comments to do what
the comment asks. You should also describe what you're
doing to make it obvious to the reader so they can
easily follow your logic in a fully reproducible way.
For any statistical tests you run, you should explain
the results in plain English and present the statistics
as you would in an APA style manuscript. For any symbols
you don't know how to use/can't use, feel free to use the name
instead (e.g. alpha)
--> 

<!-- Load the politics.csv data file. -->

First I need to load the data
```{r}
setwd("C:/Users/Schab/Desktop/Capstone Assignments/data")
politics<-read.csv("politics.csv")
```


<!-- 
Check the structure.
subject, party, testtime, minwage, and sex need to be factors.
optimismscore and income should be integer and numeric vectors.
Fix any variables that aren't right.
-->

First I need to look over the data.
```{r}
str(politics)
```

After looking at the data, subjects should be a factor and not a integer. I will explicitly tell R to change subject to a factor.  Additionally, the categorical data type (e.g. post and pre) for testime should be switched around. Therefore, I will refactor those two variables and double check my work to ensure all commands were changed correctly.

```{r}
politics$subject<-factor(politics$subject)
politics$testtime<-factor(politics$testtime, levels=c("pre", "post"))
str(politics)
```
<!-- Find the minimum, mean, and variance for posttest
optimism scores. -->

I will use the R to find the minimum, mean, and variance.

```{r}
min(politics$optimismscore[politics$testtime=="post"])
mean(politics$optimismscore[politics$testtime=="post"])
var(politics$optimismscore[politics$testtime=="post"])
min(politics$optimismscore)
mean(politics$optimismscore)
var(politics$optimismscore)
```

<!-- Create a histogram of posttest optimism scores.
Feel free to use the default ugly settings since
we didn't cover how to do this in ggplot2. If
you're a real go-getter, see if you can figure
out how to make a prettier histogram with better
labels, etc. using ggplot2. -->

For this section I will specifically command R to run only data from the optimismscores on the posttest.

```{r}
hist(politics$optimismscores[politics$testtime=="post"])
hist(optimismscores)
hist(optimismscores[politics$testime=="post"])
hist(politics$optimismscore[politics$testtime=="post"])
```

<!-- Run a test to determine whether political affiliation
is independent of one's gender.
Present the data supporting your analyses in graphical or
tabular form.
-->

Ok, first I must see if the two categorical variables are independent of one another. I can do this by calculating the frequency that individuals appear in different political groups.  Additionally, I will use a Pearson's Chi-square test of independence to test the hypothesis and determine whetherthe two variables (affiliation and gender) are independent.  I will only calculate posttest data to ensure people are not counted twice. 

```{r}
pre=politics[trues,]
post=politics[!trues,]

tab<-table(politics$party[politics$testtime=="pre"],politics$sex[politics$testtime=="pre"])
tab
mytable<-table(pre$party,pre$sex)
mytable

margin.table(mytable,1) #Row margins
margin.table(mytable,2) #Row margins
chisq.test(politics$party[politics$testime=="pre"]),politics$sex[politics$testime=="pre"]
chisq.test(pre$party,pre$sex)
```
Further, the results from the chi-test reflect the different sex and their party affiliation and how they fell into different conditions.
R(i.e squared)(2)=.72674,p=0.6953. the results are not significant can reject the null hypothesis that party affiliation is independent of sex.


<!-- Run a test to determine whether or not males and females
have different incomes.
Create a graph to display your data that includes the relevant
means and standard errors.-->

To see if there is an difference between incomes for male and female, I will run a t-test. Additionally, since males and females represent different groups of people we'll need to use an independent t-test.

```{r}
t.test(pre$[pre$sex=="male"],
      pre$income[pre$]sex=="female"],
      paired=FALSE, Var.equal=FALSE,
      alternative="two.sided",
      conf.level=.95
t.test(politics$income[politics$sex=="male"&
                         politics$testtime=="pre"],politics$income[politics$sex=="female"& politics$testtime=="pre"])
t.test(politics$income[politics$sex=="male" & politics$testtime=="pre"],politics$income[politics$sex=="female" & politics$testtime=="pre"])
```

t(62.9)=-.37,p=.71. as, you can see from the output, males and females did differ. That is, in terms of income.

Next, I will need to create the figure. I must command R I want to work with dplyr library.  I will the dplyr library to summarize my politics dataset. I will then calculate the the means and standard errors of the means for the different conditions.  Next, I will command R to use the ggplot2 library to create a nice figure.  Further, this will be the same chain of action used for any other figures I will create unless I commmand otherwise.

```{r}
library("dplyr")
temp<-politics[politics$testtime=="pre",]%>%group_by(sex)%>%summarize(means=mean(income),
              sems=sd(income)/sqrt(length(income)))
library("ggplot2")
f<-ggplot(temp,aes(x=factor(sex),y=means))+
    geom_bar(stat="identity", color="black",fill=c("deeppink","deepskyblue2"))+
    geom_errorbar(aes(ymax=means+sems, ymin=means-sems), width=.1)
f
```

This is a simple figure, but I will create a more elaborate figure with the following commands.

```{r}
f<-f+ggtitle("Incomes by Sex")+
    labs(x="Sex", y="Pretest Income Scores\n(higher=more Income")+
    scale_x_discrete(breaks=c("female","male"),labels=c("Female","Male"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(), axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())
f
```
Awesome!

<!-- See if male and females with different political affiliations
have different posttest optimism scores. Create a graph to display
your data that includes the relevant means and standard errors.-->

Ok, I will use a two-way Between Subjects ANOVA. (That is, to see if the  variables party affiliation and sex predict optimismscore independently). Further, to observe interactions between the independent variables.
```{r}
t.test(political$optimism[(politics$testtime=="pre")&
                          (politics$party=="republican")],
       politics$optimism[(politics$testtime=="post")&
                           (politics$party=="republican")],
       paired=TRUE, conf.level=1-.05/3
       
summary(aov(optimismscore~party*sex,data=politics[politics$testtime=="post",]))
```
From these results, I can conclude that optimism score is associated with political affiliation, F(2,60)=27.063,p=4.2 Income is not associated to sex, F(1,60)=0.040,p=.0843. Furthermore, there was no interaction between affiliation sex, F(2,60)=.1.213,p=.0.304.

Next, I will command R to create a plot and manually adjust the fill.  This will be done by converting the named colors to hexadecimal and using the gplots library.

```{r}
temp<-politics[politics$testtime=="post",]%>%group_by(party,sex)%>%
    summarize(means=mean(optimismscores),sems=sd(optimsimscores)/sqrt(length(optimismscores)))
library("gplots")
col1=col2hex("deeppink")
col2=col2hex("deepskyblue2")
f<-ggplot(temp, aes(x=party, y=means, fill=sex))+
    geom_bar(stat="identity",position=position_dodge())+
    scale_fill_manual(values=c(col1,col2),name="Sex",breaks=c("female","male"),labels=c("Female", "Male"))+
    theme(legend.key=element_rect(color="black"))+
    geom_errorbar(aes(ymax=means+sems, ymin=means-sems),width=.2,position=position_dodge(.9))+
    ggtitle("Optimism Scores by Sex and Political Affiliation")+
    labs(x="Political Party Affiliation",y="Income\n(thousands of dollars)")+
    scale_x_discrete(breaks=c("democrat","independent","republican"),labels=c("Democrat","Independent","Republican"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(),axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())+
    theme(legend.position=c(.2,.76))+
    theme(legend.background=element_blank())+
    theme(legend.background=element_rect(color="black"))+
    theme(legend.title=element_blank())+
    theme(legend.title=element_text(size=12))+
    theme(legend.title.align=.5)+
    theme(legend.text=element_text(size=10,face="bold"))
f
```
<!-- Create a line graph for the same data. With
error bars of course!-->
Although line graphs are created for interval or ratio scale variable.  It's clear I'm creating this graph for no other reason than pracice
```{r}
f<-ggplot(temp, aes(x=party, y=means, group=sex, color=sex))+
    geom_line(size=1)+
    geom_point(size=2)+
    scale_color_manual(values=c(col1,col2),name="Sex",breaks=c("female","male"),labels=c("Female", "Male"))+
    geom_errorbar(aes(ymax=means+sems, ymin=means-sems),width=.2)+
    ggtitle("Optimsim Score by Sex and Political Affiliation")+
    labs(x="Political Party Affiliation",y="Income\n(thousands of dollars)")+
    scale_x_discrete(breaks=c("democrat","independent","republican"),labels=c("Democrat","Independent","Republican"))+
    theme(plot.title=element_text(size=15,face="bold",vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold",vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold",vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    coord_cartesian(ylim=c(min(temp$means)-2*max(temp$sems),max(temp$means)+2*max(temp$sems)))+
    theme(panel.border=element_blank(),axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())+
    theme(legend.position=c(.2,.76))+
    theme(legend.background=element_blank())+
    theme(legend.background=element_rect(color="black"))+
    theme(legend.title=element_blank())+
    theme(legend.title=element_text(size=12))+
    theme(legend.title.align=.5)+
    theme(legend.text=element_text(size=10,face="bold"))
f
```

<!-- Run a test to see if optimism scores pre- and post
watching videos depends on sex.-->

The best course of action to determine if optimism scores before and after watching videos vary depending on sex, I must run a 2 way mixed ANOVA.  I'm using the mixed ANOVA because one or more of the variables is within-subjects and one or more of the variables is between-subjects. That is, subjects indicated their optimism before and after the videos (i.e a within-subjects factor) and because different people are affiliated with different political parties (i.e. a between-subjects factor).

```{r}
summary(aov(optimismscore~testtime*sex+Error(subject/testtime),data=politics))
```

From these results, I can conclude that optimism was related to their party affiliation. F(1,64)=0.119,p<.001, and watching videos did affect optimism, F(1,64)=41299,p<.001.  Furthermore, the results also show there was no interaction between taking the test before and after watching videos and their sex, F(1,64)=.049,p=.825.

<!-- Run a Multiple Regression Analysis to see
whether you can predict someones posttest optimism
scores from their pretest optimism scores and party
affiliation. Create a figure that shows the distributions
of pre- and posttest optimism scores and sex and create
lines that best fit the data for both sexes regardless of
whether the lines are warranted based on the statistical
analyses-->

I will command R to use Multiple Regression to predict the posttest outcome for optimism scores from income and sex.

```{r}
summary(lm(optimismscore[testtime=="post"]~optimismscore[testtime=="pre"]+party [testtime=="pre"],data=politics))
```

From these results, we would conclude that party affiliation and sex do not predict optimism scores in the pretest, R2(2)=.014,F(2,36)=.44,p=.65.  Although you typically would not create a scatter plot without significant lines that best fit. I have been instructed for practice to create one.

```{r}
f<-ggplot(politics[politics$testtime=="pre",],aes(x=income,y=optimismscore,color=sex))+
    geom_point(size=2)+
    geom_abline(intercept=60.2+2.51/2, slope=-.092,color=col1)+
    geom_abline(intercept=60.2-2.51/2, slope=-.092,color=col2)+
    scale_color_manual(values=c(col1,col2),breaks=c("female","male"),labels=c("Female","Male"))+
    ggtitle("Optimism Predicted by Sex and party")+
    labs(x="Party Affiliation",y="Optimism Score\n(Higher=More)")+
    theme(plot.title=element_text(size=15,face="bold", vjust=.5))+
    theme(axis.title.x=element_text(size=12,face="bold", vjust=-.25))+
    theme(axis.title.y=element_text(size=12,face="bold", vjust=1))+
    theme(axis.text.x=element_text(size=10,face="bold",color="black"))+
    theme(axis.text.y=element_text(size=10,face="bold",color="black"))+
    theme(panel.border=element_blank(), axis.line=element_line())+
    theme(panel.grid.major.x=element_blank())+
    theme(panel.grid.minor.x=element_blank())+
    theme(panel.grid.major.y=element_line(color="darkgrey"))+
    theme(panel.grid.minor.y=element_blank())+
    theme(legend.position=c(.83,.86))+
    theme(legend.background=element_blank())+
    theme(legend.title=element_blank())+
    theme(legend.text=element_text(size=10,face="bold"))
f
```
